Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2018-01-29 Abstract.tex

In modern programming languages, queries on in-memory collections are often more
expensive than needed.
While database queries can be readily optimized, collection queries often don't
translate trivially to databases, as modern programming languages are far more
expressive than databases, supporting both nested data and first-class functions.

Collection queries can be optimized and incrementalized by hand, but this
reduces modularity, and is often too error-prone to be feasible or to enable
maintenance of resulting programs.
In this thesis, we optimize and incrementalize such collection queries to free
programmers from such burdens.
Resulting programs are expressed in the same core language, so that they can be
subjected to other standard optimizations.

@@

To incrementalize higher-order programs by program transformation, we extend
\emph{finite differencing} \citep{Paige82FDC,Blakeley:1986:EUM,Gupta99MMV} and
develop the first approach to incrementalization by program transformation for
higher-order programs. Base programs are transformed to \emph{derivatives},
programs that transform input \emph{changes} to output changes.
We prove that our incrementalization approach is correct: We develop the theory
underlying this approach for simply-typed and untyped $\lambda$-calculus, and
discuss extensions to System F\@.

>>>>

In modern programming languages, queries on in-memory collections are often more
expensive than needed.
While database queries can be readily optimized, it is often not trivial to use
them to express collection queries which employ nested data and first-class
functions, as enabled by functional programming languages.

Collection queries can be optimized and incrementalized by hand, but this
reduces modularity, and is often too error-prone to be feasible or to enable
maintenance of resulting programs. To free programmers from such burdens,
in this thesis we study how to optimize and incrementalize such collection queries.
Resulting programs are expressed in the same core language, so that they can be
subjected to other standard optimizations.

@@

To incrementalize higher-order programs by program transformation, we extend
\emph{finite differencing} \citep{Paige82FDC,Blakeley:1986:EUM,Gupta99MMV} and
develop the first approach to incrementalization by program transformation for
higher-order programs. Base programs are transformed to \emph{derivatives},
programs that transform input \emph{changes} to output changes.
We prove that our incrementalization approach is correct: We develop the theory
underlying incrementalization for simply-typed and untyped $\lambda$-calculus, and
discuss extensions to System F\@.
